---
title: "RUVIII Simulation Report 2.0"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Preamble

This script simulates mass cytometry data and tests the effects of RUVIII normalisation on this data.
Broadly the script simulates data from the model: 

$$ Y = X\beta + W\alpha + \epsilon $$
Which is described in detail in the papers about RUV normalistion. Adding to this model we also simulate mixture normal noise for some of the parameters which causes clustering in the high dimensional space.

We then attempt to remove batch effects useing RUVIII with pseduo-replicates. For this technique we must select which pseduo replciates we are using and which markers we use as negative controls. 

Libraries

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(ggpubr)
devtools::load_all()
set.seed(44)
library(RColorBrewer)
library(distr)
# Stop the warning
distroptions(WarningArith = FALSE)
```

Data Simulation

```{r}
# PARAMETERS
ncells <- 20000 # Total number of cells
nmarkers <- 20 # Number of markers
beta_size <- 5 # Sample differnce
sigma_alpha <- 10 # Size of batch effects
sigma_epsilon <- 4 # Size of random noise
bio_param_locs <- c(1, 3, 5, 7)
mixture_size <- 20
nbioclust <- 2^(length(bio_param_locs))

# Global mean (similair overall shift to CyTOF data)
mu <- matrix(10, ncells, nmarkers)

# Sample (treatment/control) design matrix
X <- matrix(c(rep(0,ncells/2), rep(1,ncells/2)), ncells, 1)

# Coefficients of interest
beta <- matrix(0, 1, nmarkers)
beta[bio_param_locs] <- rep(beta_size, length(bio_param_locs))

# Random error
epsilon <- matrix(rnorm(ncells * nmarkers, sd = sigma_epsilon), ncells, nmarkers)

# RUVIII model
Y <-  mu + X %*% beta

# PARAMETER
# Mixing coefficients
mix_coeff <- c(1/2, 1/2)
multnorm <- UnivarMixingDistribution(Norm(-mixture_size, 1), Norm(mixture_size, 1), 
                                     mixCoeff = mix_coeff)
my_rmultnorm <- r(multnorm)

Z <- matrix(0, ncol = nmarkers, nrow = ncells)

for(i in 1:ncol(Y)){
  if(i %in% bio_param_locs){
    bio <- my_rmultnorm(ncells)
    Y[,i] <- Y[,i] + bio
    Z[,i] <- ifelse(bio > 0, 1, 0)
  } else {
    # Do nothing
  }
}

bio_factors <- lapply(bio_param_locs, function(x) factor(Z[, x]))

int <- interaction(bio_factors)

W_vec <- numeric(ncells)

# PARAMETER
# Proportion of cells with batch effect in a given cluster
# NB: Must have length `nbioclust`
#props <- c(0.3, 0.7, 0.2, 0.8)
props <- rep(0.5, nbioclust)

# PARAMETER
# Correlation between X and W
# Here W is constructed so that it is uncorrelated with X
for(i in 1:length(levels(int))){
  ind <- which(int == levels(int)[i])
  W_vec[ind] <- rbinom(length(ind), 1, props[i])
}

# Unwanted variation design matrix
# Note: This is UNCORRELATED with X
W <- matrix(W_vec, ncells, 1)

#W <- matrix(rbinom(ncells, 1, 0.5), ncells, 1)

# Coeffcients of unwanted variation
alpha <- matrix(rnorm(nmarkers, sd = sigma_alpha), 1, nmarkers)

Y <- Y + W %*% alpha

Y <- Y + epsilon
```

Plot the data

```{r}
# PARAMETER
# Size of the subset
N <- 1000
samp <- sample(1:20000, N)

plot_data <- tibble(x = Y[samp,3], 
                    y = Y[samp,1],
                    sample = factor(as.vector(X[samp, ])),
                    batch = factor(as.vector(W[samp, ])),
                    bio = int[samp])

# PARAMETER 
# Use PCA for dimension reduction
PCA <- FALSE
if(PCA){
    pca <- prcomp(plot_data[, bio_param_locs], scale. = TRUE)
    plot_data$x <- pca$x[,1]
    plot_data$y <- pca$x[,2]
}

# PARAMETER 
# Use t-SNE for dimension reduction
# NB: tSNE only recommended if dim > 2
TSNE <- TRUE
if(TSNE){
  # Code to correct bug in the cytofkit package.
  # A passed matrix must be named...
  tsne_data <- Y[samp, bio_param_locs]
  colnames(tsne_data) <- as.character(1:ncol(tsne_data))
  # Run t-SNE
  tsne <- cytofkit::cytof_dimReduction(tsne_data, method = "tsne")
  plot_data$x <- tsne[,1]
  plot_data$y <- tsne[,2]
}

sample_plot <- ggplot(plot_data, aes(x, y, col = sample)) + 
                  geom_point() + 
                  theme_bw() + 
                  labs(x = "", y = "", col = "Sample") + 
                  scale_colour_manual(values=c("steelblue", "green4"))

bio_plot <- ggplot(plot_data, aes(x, y, col = bio)) + 
                geom_point() + 
                theme_bw() + 
                labs(x = "", y = "", col = "Biology")
            
batch_plot <- ggplot(plot_data, aes(x, y, col = batch)) + 
                 geom_point() + 
                 theme_bw() + 
                 labs(x = "", y = "", col = "Batch") + 
                 scale_colour_manual(values=c("red3", "royalblue3"))

sample_plot
bio_plot
batch_plot
```

```{r}
boxplot_data <- data.frame(Y, 
                           sample = factor(as.vector(X)),
                           batch = factor(as.vector(W)),
                           bio = int)

boxplot_data <- gather(boxplot_data, key = "marker", value = "intensity",
                       X1:X20, -sample, -batch, -bio)


sample_boxplot <- ggplot(boxplot_data, aes(marker, intensity, fill = sample)) + 
                      geom_boxplot() + 
                      theme_bw() + 
                      labs(x = "", y = "", col = "Sample") + 
                      scale_fill_manual(values=c("steelblue", "green4"))

bio_boxplot <- ggplot(boxplot_data, aes(marker, intensity, fill = bio)) + 
                  geom_boxplot() + 
                  theme_bw() + 
                  labs(x = "", y = "", col = "Biology") + 
                  facet_wrap(~bio)

batch_boxplot <- ggplot(boxplot_data, aes(marker, intensity, fill = batch)) + 
                   geom_boxplot() + 
                   theme_bw() + 
                   labs(x = "", y = "", col = "Batch") + 
                   scale_fill_manual(values=c("red3", "royalblue3"))

sample_boxplot
bio_boxplot
batch_boxplot
```


```{r}
# PARAMETER
# Which what sort of pseudo-replicates we are using 

# Factor which includes all of the 4 biological clusters
bio_int <- int

# Factor which includes all of the biological clusters and the sample difference
bio_factors[[length(bio_param_locs) + 1]] <- factor(X[,1])
bio_samp_int <- interaction(bio_factors)

# Use the ruv package to convert the factor to a replicate matrix
M <- ruv::replicate.matrix(bio_samp_int)

#M <- cbind(M[,1], M[,2], M[,3], !(M[,1] | M[,2] | M[,3]))
#M <- cbind(W, !W)


# PARAMETER
# What fraction (0,1) of the simulated samples should we make random

rand_frac <- 0
peturb_ind <- sample(1:nrow(M), rand_frac * nrow(M))
for(i in peturb_ind){
  M[i, ] <- sample(0:1, ncol(M), replace = TRUE)
}

# Standardise the cell data
# Could refactor RUVIII wrappers to make this easier
raw_Y <- as.matrix(Y)

col_means <- colMeans(raw_Y)
col_sds <- apply(raw_Y, 2, function(x) sd(x))

for(i in 1:ncol(raw_Y)){
  raw_Y[,i] <- (raw_Y[,i] - col_means[i])/col_sds[i]
}

# PARAMETER
# What negative controls are we using

# The negative contols will taken to be all the markers
# except for `exclude_neg_controls_indices` 
exclude_neg_controls <- c(1,3,5)
controls <- c(1:nmarkers)[-exclude_neg_controls]

RUVIIout <- fastRUVIII(Y = Y, M, ctl = controls, k = 1)
norm_Y <- RUVIIout$newY
alpha <- RUVIIout$fullalpha

for(i in 1:ncol(norm_Y)){
  norm_Y[,i] <- norm_Y[,i]*col_sds[i] + col_means[i]
}
```

```{r}
norm_plot_data <- tibble(x = norm_Y[samp,3], 
                         y = norm_Y[samp,1],
                         sample = factor(as.vector(X[samp, ])),
                         batch = factor(as.vector(W[samp, ])),
                         bio = int[samp])

# PARAMETER 
# Use t-SNE for dimension reduction
TSNE <- TRUE
if(TSNE){
  # Code to correct bug in the cytofkit package.
  # A passed matrix must be named...
  tsne_data <- norm_Y[samp, bio_param_locs]
  colnames(tsne_data) <- as.character(1:ncol(tsne_data))
  # Run t-SNE
  tsne <- cytofkit::cytof_dimReduction(tsne_data, method = "tsne")
  norm_plot_data$x <- tsne[,1]
  norm_plot_data$y <- tsne[,2]
}

sample_plot <- ggplot(norm_plot_data, aes(x, y, col = sample)) + 
                      geom_point() + 
                      theme_bw() + 
                      labs(x = "", y = "", col = "Sample") + 
                      scale_colour_manual(values=c("steelblue", "green4"))

bio_plot <- ggplot(norm_plot_data, aes(x, y, col = bio)) + 
                    geom_point() + 
                    theme_bw() + 
                    labs(x = "", y = "", col = "Biology")

batch_plot <- ggplot(norm_plot_data, aes(x, y, col = batch)) + 
                     geom_point() + 
                     theme_bw() + 
                     labs(x = "", y = "", col = "Batch") + 
                     scale_colour_manual(values=c("red3", "royalblue3"))

sample_plot
bio_plot
batch_plot
```

```{r}
boxplot_data <- data.frame(norm_Y, 
                           sample = factor(as.vector(X)),
                           batch = factor(as.vector(W)),
                           bio = int)

boxplot_data <- gather(boxplot_data, key = "marker", value = "intensity",
                       X1:X20, -sample, -batch, -bio)


sample_boxplot <- ggplot(boxplot_data, aes(marker, intensity, fill = sample)) + 
                      geom_boxplot() + 
                      theme_bw() + 
                      labs(x = "", y = "", col = "Sample") + 
                      scale_fill_manual(values=c("steelblue", "green4"))

bio_boxplot <- ggplot(boxplot_data, aes(marker, intensity, fill = bio)) + 
                  geom_boxplot() + 
                  theme_bw() + 
                  labs(x = "", y = "", col = "Biology") + 
                  facet_wrap(~bio)

batch_boxplot <- ggplot(boxplot_data, aes(marker, intensity, fill = batch)) + 
                   geom_boxplot() + 
                   theme_bw() + 
                   labs(x = "", y = "", col = "Batch") + 
                   scale_fill_manual(values=c("red3", "royalblue3"))

sample_boxplot
bio_boxplot
batch_boxplot
```
